Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz Avila"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: Julius Center for Health Sciences and Primary Care
    address: |
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands   
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
    affiliation2: |
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, Amsterdam, The Netherlands

@@

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random effect       & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random coefficients \\
                    & and heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. [TODO: add stratification.] \\
\hline
\end{tabular}
\end{table}

@@

\begin{table}[tb]
\caption{Notation}
\label{tab:mod}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
& [TODO: explain \pkg{lme4} notation here] \\
\hline
\end{tabular}
\end{table}

@@

Impute the data:
```{r eval=FALSE}
imp <- mice(obesity, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```

@@

>>>>

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: Julius Center for Health Sciences and Primary Care
    address: |
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands   
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
    affiliation2: |
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, Amsterdam, The Netherlands

@@

\begin{table}[tb]
\caption{Concepts in multilevel methods}
\label{tab:clus}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
Sample unit         & Units of the population from which measurements are taken in a sample.\\                           & These should be distinct and not overlapping, e.g., patients, students.\\ 
Hierarchical levels & Data are grouped into clusters at different levels. A three-level \\                               & hierarchical model could occur by collecting observations of math scores\\                         & (lower level) which can be measured over the whole school year for each\\                          & student, and then grouped by student (middle level) and these in turn can\\                        & be aggregated into classrooms (high level).\\
Fixed effect        & Here we assume that the values of an independent variable are fixed, i.e., \\ 
  & the values observed in the study are representative of all values in the \\                        & population we are interested in. In the case of a fixed factor variable, for \\                     & example treatment type, we are only concerned with assessing the difference \\ 
                & in the dependent variable y e.g., blood pressure between treatments A and B.\\                     & There may be additional treatments, but based on the objective of the study,\\                     & which in this case is the equivalence effect of A and B, the other treatment\\                     & types are irrelevant and therefore the treatment type is included as a fixed \\                     & effect with only the A and B levels in the hierarchical model. \\
Random effect       & The values of an independent variable are assumed to be randomly drawn from \\                     & a population with more values. In a random factor, the levels observed in the\\                     & study represent a sample of all possible levels, e.g., hospitals, classrooms.\\                     & For example, in a study to predict the blood pressure of patients on \\  
& admission we might select only certain hospitals that are representative of \\                     & the entire population of hospitals. Here we are not interested in measuring \\ 
& the difference of y between individual hospitals, but rather the variation of\\                     & y between hospitals, to create a model that can be generalized to the entire\\                     & population of hospitals. \\

ICC                 & The variability due to clustering is often measured by means of the \\
                    & intraclass coefficient (ICC). The ICC can be seen as the percentage \\
                    & of variance that can be attributed to the cluster-level, where a high \\
                    & ICC would indicate that a lot of variability is due to the cluster \\
                    & structure. \\
Random effect       & Multilevel models typically accommodate for variability by including \\
                    & a separate group mean for each cluster. In addition to random \\
                    & intercepts, multilevel models can also include random coefficients \\
                    & and heterogeneous residual error variances across clusters [see e.g. \\
                    & @gelm06, @hox17 and @jong21]. [TODO: add stratification.] \\
\hline
\end{tabular}
\end{table}

@@

```{=tex}
\begin{table}[tb]
\caption{Notation}
\label{tab:mod}
\centering
\begin{tabular}{lll}
\hline
\textbf{Formula lme4} & \textbf{Details}   \\
\hline
y$\sim$ x1 + (1 | g1)                                       & Fixed x1 predictor with random intercept  \\
                                                         & varying among g\\
y$\sim$x1*x2+ (1| g1)                                      & Interactions of x1 and x2 only in fixed effect\\
y$\sim$x1*x2+ (x2| g1)                                     & Interactions of x1 and x2 only in fixed effect\\
                                                         & with slope of x2 randomly varying among g1\\
y$\sim$x1*x2+ (x1*x2| g1)                                  & variance-covariance matrix estimated only with \\
                                                         & the variance terms of intercept, slope of x1, \\
                                                         & slope of x2 and interaction x1*x2\\
y$\sim$x1*x2+ (x1| g1)+ (x2| g1)                           & variance-covariance matrix estimated separately,\\
                                                         & i.e, one for intercept and x1 and another for \\
                                                         & intercept and x2\\
y$\sim$x1 + (x1 | g1) or 1 + x1 + (1 + x1 | g1)            & Fixed x1 with correlated random intercept and \\
                                                         & random slope of x\\
y$\sim$x1 + (x1 || g1) or 1 + x1 + (1 | g1) + (0 + x1 | g1) & Fixed x1 with uncorrelated random intercept\\
                                                         & and random slope of x1\\  
y$\sim$(1 | g1) + (1 | g2)                                 & Random intercept varying among g1 and among g2\\
y$\sim$(1 | g1/g2) or ~(1|g1)+(1|g1:g2)                    & Random intercept varying among g1 and g2 
                                                         & withing g1\\

& [TODO: explain \pkg{lme4} notation here
https://datalorax.github.io/equatiomatic/articles/lmer4-lmer.html
https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf
] \\
\hline
\end{tabular}
\end{table}
```

@@

```{r eval=FALSE}
imp <- mice(obesity, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```

@@

-  I don't know if in your article you cover something about model complexity, for example sometimes I have to switch from 2l. methods to 1l. methods just because the model didn't converge.. this is due to the considered imputation model is very complex regarding the amount of information counted... I know that a solution for an imputation model with many predictors is to check correlation plots as you did or use quickpred().. (maybe you can add this somewhere after the correlation plots).....But as for cluster specification, I don't know if besides plots of distribution per cluster there is something else can be done to see if i have to use 1l. or 2l. for a given variable, also I have no idea.. how to test which is better between 2l.norm or 2l.2stage.norm.
